---
output: html_document
---

```{r setup, include=FALSE}
# general packages
library(tidyverse)
library(conflicted)
library(magrittr)

# plotting packages
library(viridis)
library(wesanderson)
library(UpSetR)

# resolve package conflicts
filter <- dplyr::filter

knitr::opts_chunk$set(echo = TRUE)
```

## Figure 3: Linear Model

Code for Figure 3 for the paper The Microbiome Modulates DNA Methylation in Mouse Colonic Mucosa

### Wrangle Data

#### Read Data

Read in the linear model results table. For code used to run the test, see `../../lm/lm.Rmd`

```{r}
read_tsv('../../lm/lm.tsv') -> lm
```

#### Wrangle Data

For **volcano plots**

```{r}
lm %>%
# log the qvalues for easier volcano plotting
  mutate(logqvalue = -log10(qvalue), 
# annotate the direction of change to count sites for volcano plot annotation
         change = ifelse(estimate < 0, 'hypo', 'hyper'),
# add a categorical column for easy volcano plot coloring
         volc_color = case_when((qvalue < 0.05 & abs(estimate) >= 5) & 
                                  term == 'inflammation1' ~ 'sig_inf',
                                (qvalue < 0.05 & abs(estimate) >= 5) & 
                                  term == 'microbiome1' ~ 'sig_mic',
                                (qvalue < 0.05 & abs(estimate) >= 5) & 
                                  term == 'drug1' ~ 'sig_drug',
                                TRUE ~ 'notsig'),
# add a categorical column for better facet names
         facets = case_when(term == 'inflammation1' ~ 'IL10KO',
                            term == 'microbiome1' ~ 'Microbiome',
                            term == 'drug1' ~ 'Azoxymethane')) -> lm_volc
```

For **UpSet plots**. The UpSet plot needs binary columns, 0/1, indicating whether the sma site is significant under that condition

```{r}
### hypOmethylated sites
lm_volc %>%
  filter(abs(estimate) >= 5) %>%
  mutate(term_change = case_when(term == 'microbiome1' & 
                                   estimate <= -5 ~ 'Microbiome_hypo',
                                 term == 'inflammation1' & 
                                   estimate <= -5 ~ 'IL10KO_hypo',
                                 term == 'drug1' & 
                                   estimate <= -5 ~ 'Azoxymethane_hypo',
                                 TRUE ~ 'hyper')) %>%
  filter(term_change != 'hyper') %>%
  select(mm9_SmaI_ID, term_change, qvalue) %>%
  spread(term_change, qvalue) %>%
  replace(is.na(.), 1) %>%
  mutate(Azoxymethane = ifelse(Azoxymethane_hypo < 0.05, 1, 0),
         IL10KO = ifelse(IL10KO_hypo < 0.05, 1, 0),
         Microbiome = ifelse(Microbiome_hypo < 0.05, 1, 0)) %>%
  as.data.frame() -> lm_upset_hypo
#write_tsv(lm_upset_hypo, 'lm_upset_hypo.tsv')

### hypERmethylated sites
lm_volc %>%
  filter(abs(estimate) >= 5) %>%
  mutate(term_change = case_when(term == 'microbiome1' & 
                                   estimate >= 5 ~ 'Microbiome_hyper',
                                 term == 'inflammation1' & 
                                   estimate >= 5 ~ 'IL10KO_hyper',
                                 term == 'drug1' & 
                                   estimate >= 5 ~ 'Azoxymethane_hyper',
                                 TRUE ~ 'hypo')) %>%
  filter(term_change != 'hypo') %>%
  select(mm9_SmaI_ID, term_change, qvalue) %>%
  spread(term_change, qvalue) %>%
  replace(is.na(.), 1) %>%
  mutate(Azoxymethane = ifelse(Azoxymethane_hyper < 0.05, 1, 0),
         IL10KO = ifelse(IL10KO_hyper < 0.05, 1, 0),
         Microbiome = ifelse(Microbiome_hyper < 0.05, 1, 0),) %>%
  as.data.frame() -> lm_upset_hyper
#write_tsv(lm_upset_hyper, 'lm_upset_hyper.tsv')
```

#### Volcano Plots

```{r, fig.width = 8, fig.height = 6}
### volcano plot annotation
lm_volc %>% 
  group_by(facets, change, volc_color) %>% 
  count() %>% 
  ungroup() %>%
  filter(volc_color != 'notsig') %>%
  mutate(perc = round(((n / 12566) * 100), 1),
         estimate = rep(c(20, -20), 3), logqvalue = 12, 
         term = rep(c('Azoxymethane', 'IL10KO', 'Microbiome'), each = 2),
         sign = '%') %>%
  unite(anno, n, perc, sep = ', ') %>%
  unite(anno, anno, sign, sep = '') -> lm_volc_counts

### plot
ggplot(lm_volc, aes(x = estimate, y = logqvalue)) + 
  geom_point(aes(color = volc_color), size = 1) + 
  scale_color_manual(values = c('gray60', 'darkorange2', 
                                'firebrick3', 'deepskyblue3')) +
  geom_hline(yintercept = -log10(0.05), 
             linetype = 'dashed', 
             color = 'gray30') +
  geom_vline(xintercept = c(-5, 5), 
             linetype = 'dashed', 
             color = 'gray30') +
  geom_text(data = lm_volc_counts, aes(label = anno), size = 5) +
  labs(x = 'Sma Site Linear Model Slope', 
       y = '-Log10 QValue',
       title = 'A') +
  facet_wrap(. ~ facets) +
  theme_classic(base_size = 20) +
  theme(legend.position = 'none')

### save
ggsave('lm_slope_volc.png', width = 8, height = 6)
ggsave('lm_slope_volc.pdf', width = 8, height = 6)
ggsave('lm_slope_volc.tiff', width = 8, height = 6)
```

#### UpSet Plots

UpSet plot for hypomethylated CpG sites.

```{r, fig.width = 8, fig.height = 6}
### significant hypOmethylated sites
# as png
png('lm_upset_hypo.png')
upset(lm_upset_hypo, 
      sets = c('Microbiome', 'IL10KO', 'Azoxymethane'),
      keep.order = T,
      order.by = 'degree',
      main.bar.color = 'dodgerblue',
      mainbar.y.label = '',
      point.size = 4,
      text.scale = c(1, 2, 1, 1.75, 2, 3))
dev.off()

# as pdf
pdf('lm_upset_hypo.pdf', onefile = FALSE)
upset(lm_upset_hypo, 
      sets = c('Microbiome', 'IL10KO', 'Azoxymethane'),
      keep.order = T,
      order.by = 'degree',
      main.bar.color = 'dodgerblue',
      mainbar.y.label = '',
      point.size = 4,
      text.scale = c(1, 2, 1, 1.75, 2, 3))
dev.off()

# as tiff
tiff('lm_upset_hypo.tiff')
upset(lm_upset_hypo, 
      sets = c('Microbiome', 'IL10KO', 'Azoxymethane'),
      keep.order = T,
      order.by = 'degree',
      main.bar.color = 'dodgerblue',
      mainbar.y.label = '',
      point.size = 4,
      text.scale = c(1, 2, 1, 1.75, 2, 3))
dev.off()
```

UpSet plot for hypermethylated CpG sites

```{r, fig.width = 8, fig.height = 6}
### significant hypERmethylated sites
# as png
png('lm_upset_hyper.png')
upset(lm_upset_hyper, 
      sets = c('Microbiome', 'IL10KO', 'Azoxymethane'),
      keep.order = T,
      order.by = 'degree',
      main.bar.color = 'firebrick3',
      mainbar.y.label = '',
      point.size = 4,
      text.scale = c(1, 2, 1, 1.75, 2, 3))
dev.off()

# as pdf
pdf('lm_upset_hyper.pdf', onefile = FALSE)
upset(lm_upset_hyper, 
      sets = c('Microbiome', 'IL10KO', 'Azoxymethane'),
      keep.order = T,
      order.by = 'degree',
      main.bar.color = 'firebrick3',
      mainbar.y.label = '',
      point.size = 4,
      text.scale = c(1, 2, 1, 1.75, 2, 3))
dev.off()

# as tiff
tiff('lm_upset_hyper.tiff')
upset(lm_upset_hyper, 
      sets = c('Microbiome', 'IL10KO', 'Azoxymethane'),
      keep.order = T,
      order.by = 'degree',
      main.bar.color = 'firebrick3',
      mainbar.y.label = '',
      point.size = 4,
      text.scale = c(1, 2, 1, 1.75, 2, 3))
dev.off()
```

### sessionInfo

```{r}
sessionInfo()
```

<br><br>
